---
title: "Modeling SSC Reductions from maxABC Using Risk Table Scores"
author: ""
date: today
bibliography: Thompson_link_ABC_reduction.bib
format:
  html:
    toc: true
    toc-depth: 3
    lightbox: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    self-contained: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(betareg)
library(brms)
library(knitr)
library(patchwork)

theme_set(theme_minimal(base_size = 13))
options(mc.cores = min(2, parallel::detectCores()))
```

## Summary

This report looks at whether groundfish risk table scores help explain when
the Scientific and Statistical Committee (SSC) recommends setting ABC below
the maximum permissible ABC. In simple terms, it asks whether higher or more
widespread concerns in the risk table tend to line up with larger reductions.

The answer is yes, but only partly. Stocks with more elevated concerns often
do receive larger reductions, and population dynamics concerns appear to be
especially informative. At the same time, the relationship is not tight enough
to treat the risk table as a formula. Similar score patterns can still lead to
different reductions because the SSC also considers stock-specific context,
model behavior, and management judgment.

The results should therefore be read as descriptive rather than prescriptive.
They summarize how risk table scores and SSC reductions have lined up in the
historical record, but they do not provide a mechanical rule for future
decisions.

## Introduction

The North Pacific Fishery Management Council's Scientific and Statistical
Committee (SSC) may set ABCs below the maximum permissible ABC (maxABC)
when the assessment risk table identifies elevated concerns. The risk table
scores four categories—Assessment-related (A), Population dynamics (P),
Ecosystem (E), and Fishery-informed (F)—on a scale where scores >1 indicate
concern.

This analysis uses the *Groundfish Risk Table Report: 2024 Update*, presented
under agenda item C1 at the December 2024 Council meeting, as its source
document for the historical risk-table record [@ShotwellBryan2024RiskTable].
That report collated risk table information for Alaska groundfish stocks in
the Bering Sea and Aleutian Islands (BSAI) and the Gulf of Alaska (GOA) from
2018 onward and framed the summary as a living reference for stock assessment
authors and reviewers.

The report emphasizes that risk table implementation changed substantially
over the 2018-2024 period [@ShotwellBryan2024RiskTable]. The 2018 trial year
included only a small number of completed tables and only three categories:
assessment, population dynamics, and ecosystem. Fishery performance was added
in 2019. In response to SSC direction, the scoring system was later reduced
from four levels to three beginning in 2023, and the category descriptions
were revised again for 2024 to better cover the intended range of conditions.
Those shifts make direct comparisons across years imperfect, especially when
contrasting the 2018 trial year and the reduced-level years from 2023 onward
with the 2019-2022 period.

The AFSC working group summarized annual category scores, the maximum score
across categories, author-proposed reductions, SSC-recommended reductions,
methods used to derive reductions, and cases where author, Plan Team, and SSC
recommendations differed [@ShotwellBryan2024RiskTable]. Across 2018-2024, the
report identified 136 completed risk tables: 77 with at least one elevated
score, 59 without elevated scores, 25 with an author recommendation to reduce
from maximum ABC, and 28 with a final reduction following Plan Team and SSC
review. The relatively small number of reductions compared with the number of
risk tables is central to the interpretation here: risk table scores are
informative, but they are not determinative on their own.

This document narrows that broader report into a modeling exercise. It focuses
on the relationship between risk table scores and the magnitude of the SSC's
recommended reduction from maxABC, and evaluates several candidate models for
describing those reductions.

## Background

### How the risk table framework evolved

Several design changes matter for interpretation of any time series built from
these scores:

1. In 2018, risk tables were used on a limited trial basis and included only
   assessment, population dynamics, and ecosystem categories.
2. In 2019, fishery performance was added as a fourth category.
3. In 2023, the four scoring levels were reduced to three: normal, increased,
   and extreme. Under that reduced system, current levels 2 and 3 are roughly
   comparable to the earlier levels 3 and 4.
4. In 2024, the text defining the categories was updated after the SSC asked
   for descriptions that better covered the full intended range of conditions.

These changes mean the raw scores are not fully stationary through time. Any
model fit to the combined record should therefore be treated as descriptive of
historical SSC behavior rather than as a stable calibration of a fixed scoring
system.

### How reductions have been derived in practice

The 2024 summary report documented several approaches authors and reviewers
have used when moving from elevated scores to a reduction from maxABC. Those
approaches include:

1. Deferring to the SSC to specify a reduction if one was judged warranted.
2. Applying tier-based advice developed for one stock status category to
   another, such as Tier 3-style advice used for Tier 1 stocks.
3. Using prior projection results when a previous scenario better captured the
   concern driving the reduction.
4. Applying probability-based methods such as the Thompson framework for
   translating elevated risk into a buffer below maxABC.
5. Selecting an ABC level intended to keep the stock above a biological
   threshold such as B20%.
6. Using an alternative model configuration with better retrospective
   performance or lower OFL and ABC values.
7. Taking a stair-step approach between the current-year estimate and a
   previous projection.
8. Adjusting recent recruitment assumptions or the reference period used in
   projections.

That diversity in methods is important for interpreting the results below. The
same nominal risk score can map to different reductions depending on stock
status, model structure, and the rationale judged most relevant by the SSC.

## Data

The data below come from SSC-recommended reductions only (not Plan Team or
author recommendations). Rollover years and records without a quantified
reduction are excluded. Blank years in the broader report generally correspond
to off-cycle assessments with no new risk table, although a reduction may
still carry over from an earlier decision. The analysis dataset below is
therefore a filtered subset of the full living summary rather than a complete
reproduction of every annual stock entry.

@tbl-data is the core analytical dataset used throughout the document.
It provides the stock-year observations, the category-level elevated scores
used as predictors, and the SSC reduction used as the response variable. The
derived columns summarize two distinct ideas that recur in the figures and
models below: overall severity through the total score, and breadth of concern
through the number of elevated categories.

```{r}
#| label: tbl-data
#| tbl-cap: "Risk table scores and SSC-recommended percent reductions from maxABC. Scores of 0 indicate the category was not elevated (score ≤ 1)."

dat <- tribble(
  ~stock,                            ~year, ~A, ~P, ~E, ~F, ~pct_red,
  "EBS Pollock",                      2018,  0,  2,  2,  0,  30,
  "EBS Pollock",                      2019,  0,  2,  2,  2,  43,
  "EBS Pollock",                      2020,  0,  0,  2,  2,  30,
  "EBS Pollock",                      2021,  2,  0,  2,  2,  11,
  "EBS Pollock",                      2022,  2,  0,  0,  0,  43,
  "EBS Pollock",                      2023,  0,  0,  2,  0,  18,
  "EBS Pacific cod",                  2019,  0,  0,  2,  0,   3,
  "AI Pacific cod",                   2023,  0,  2,  2,  0,  10,
  "AK Sablefish",                     2018,  2,  4,  2,  0,  45,
  "AK Sablefish",                     2019,  2,  3,  2,  3,  50,
  "AK Sablefish",                     2020,  3,  3,  2,  3,  44,
  "BSAI Yellowfin sole",              2021,  0,  2,  2,  0,  24,
  "BSAI Greenland turbot",            2024,  3,  2,  2,  0,  25,
  "BSAI Northern rock sole",          2022,  3,  0,  0,  0,  23,
  "BSAI Blackspotted/Rougheye RF",   2022,  3,  2,  0,  2,  11,
  "BSAI Sharks",                      2022,  3,  2,  0,  0,  13,
  "GOA Pollock",                      2023,  2,  0,  0,  0,  18,
  "GOA Dusky Rockfish",               2020,  2,  0,  0,  0,  24,
  "GOA Rougheye/Blackspotted RF",     2023,  2,  2,  0,  0,  20,
  "GOA Demersal shelf rockfish",      2021,  2,  2,  0,  2,  22,
  "GOA Demersal shelf rockfish",      2022,  2,  2,  0,  2,  15
)

dat <- dat |>
  mutate(
    total_score  = A + P + E + F,
    n_elevated   = (A > 0) + (P > 0) + (E > 0) + (F > 0),
    pct_red_prop = pct_red / 100,
    across(
      c(A, P, E, F),
      ~ case_when(
        year >= 2023 & .x == 2 ~ 3,
        year >= 2023 & .x == 3 ~ 4,
        TRUE ~ .x
      ),
      .names = "{.col}_harmonized"
    ),
    total_score_harmonized = A_harmonized + P_harmonized +
      E_harmonized + F_harmonized,
    n_elevated_harmonized = (A_harmonized > 0) + (P_harmonized > 0) +
      (E_harmonized > 0) + (F_harmonized > 0)
  )

dat |>
  select(stock, year, A, P, E, F, total_score, n_elevated, pct_red) |>
  kable(col.names = c("Stock", "Year", "A", "P", "E", "F",
                       "Total", "# Elevated", "% Reduction"))
```

For model sensitivity, scores from 2023 onward are also harmonized to the
earlier four-level scale. Under that harmonization, the current level 2
("increased") is treated as approximately equivalent to the earlier level 3,
and the current level 3 ("extreme") is treated as approximately equivalent to
the earlier level 4. Scores of 0 remain non-elevated placeholders in the
analysis dataset.

## Modeling historical patterns of ABC reductions below the maximum permissible

Before the risk table was formalized, Thompson prepared a working paper for
the SSC that asked a related question: could historical BSAI and GOA
Groundfish Plan Team recommendations to set ABC below the maximum permissible
level be reverse-engineered into transparent decision rules?
[@Thompson2019ABCReductions]. The analysis covered 2003-2017 and identified
76 instances in which the Plan Teams recommended ABC below `maxABC`, or about
five instances per year.

The raw historical record was qualitative. For each reduction, the Plan Teams
usually provided one or more reasons, and Thompson condensed 122 unique stated
reasons into 66 initial binary variables. Because many reasons occurred
together and made the design matrix singular, the variables were progressively
combined into 52 and then 32 final "Round 3" variables. The response was the
proportional reduction from `maxABC`, transformed with a symmetric logit-style
mapping,

$$
\log\left(\frac{1 + y_i}{1 - y_i}\right),
$$

where $y_i$ is the proportional ABC reduction for instance $i$.

The first model used ordinary least squares on the transformed response, but
some coefficients were negative. Thompson then fit a constrained version with
nonnegative coefficients so that any applicable reason would imply either no
reduction or a downward adjustment from `maxABC`, never an upward adjustment.
Both models fit the historical reductions similarly well after transformation
back to the natural scale, with $R^2$ values of about 0.83 for the
unconstrained model and 0.82 for the constrained model
[@Thompson2019ABCReductions].

The constrained model is useful here because it shows an earlier attempt to
translate qualitative concerns into a quantitative ABC buffer. Several
variables implied little or no univariate reduction once paired historical
patterns were accounted for, while others implied substantial reductions when
considered alone. Examples from the constrained model are shown in
@tbl-thompson-2019-selected.

```{r}
#| label: tbl-thompson-2019-selected
#| tbl-cap: "Selected constrained-model coefficients and implied univariate ABC reductions from Thompson's historical ABC reduction working paper."

tribble(
  ~reason, ~beta, ~univariate_reduction,
  "Projected biomass decline", 0.21, 0.10,
  "Long-term poor recruitment", 0.22, 0.11,
  "Dorn's buffer", 0.23, 0.11,
  "Uncertain recruitment estimates, including recent large estimates", 0.40, 0.20,
  "Model uncertainty", 0.43, 0.21,
  "Uncertainty surrounding M and Q", 0.60, 0.29,
  "Low or long-term declining biomass", 0.82, 0.39,
  "Large survey CV", 0.91, 0.43,
  "No new survey", 1.81, 0.72
) |>
  mutate(
    beta = round(beta, 2),
    `Univariate reduction (%)` = round(100 * univariate_reduction)
  ) |>
  select(
    Reason = reason,
    `Constrained beta` = beta,
    `Univariate reduction (%)`
  ) |>
  kable()
```

That working paper also flagged important limitations that carry over to any
analysis of risk-table reductions. The historical reasons were not collected
from a prospective checklist; they were usually articulated after an author or
Plan Team had already become concerned that a reduction might be warranted.
The variable consolidation was therefore partly subjective, and a zero
coefficient did not necessarily mean a concern was unimportant. It meant only
that the concern did not add explanatory power beyond the other concerns with
which it had historically appeared [@Thompson2019ABCReductions].

The current analysis differs in two ways. First, it uses the formal risk table
scores rather than the older ad hoc reason codes. Second, it focuses on final
SSC reductions rather than Plan Team recommendations. The connection is that
both exercises treat ABC reductions as historically observable decisions that
can be modeled descriptively, while recognizing that the fitted relationships
are not automatic decision rules.

## Thompson's link between risk and ABC reductions

Thompson's Appendix 2.6 in the 2020 EBS Pacific cod assessment
[@Thompson2020EBSPcod] proposed a direct probability link between the risk
table and the reduction from maximum permissible ABC. The currency is the
probability that `maxABC` exceeds the true-but-unknown OFL. The usual
assessment uncertainty gives a baseline probability, $P_0$, and the risk table
adds extra probability from concerns that are not already represented in the
assessment model.

For each risk category $j$, Thompson first maps the assigned risk level and an
within-level fraction, $\Delta_j$, to a conditional probability
$P_{\mathrm{cond},j}$. With four risk levels, the probability intervals are:
level 1 from $P_0$ to 0.5, level 2 from 0.5 to 2/3, level 3 from 2/3 to 5/6,
and level 4 from 5/6 to 1.0. The marginal category probabilities are then
combined as

$$
P = 1 - (1 - P_0) \prod_j (1 - p_j),
$$

where $P$ is the joint probability that `maxABC` exceeds true OFL after
accounting for all risk categories. The probability-equivalent value of
`maxABC` is then

$$
maxABC_{PE} = cdf^{-1}(P).
$$

The final buffer follows from the ratio $OFL / maxABC_{PE}$. If
$maxABC_{PE}$ is greater than OFL, ABC is reduced by that ratio; otherwise ABC
remains at `maxABC`.

Because the risk table changed through time, the applied calculation below
uses three risk levels throughout, matching the current normal, increased, and
extreme scoring. The number of categories remains year-specific: the 2018
trial-year calculations use three categories because the fishery-informed
category had not yet been added, while later calculations use four categories.

@fig-thompson-heatmap gives a first illustration of the mechanism by reducing
the four-category risk table to two simple dimensions: the number of elevated
categories and the common three-level risk score assigned to those elevated
categories. The
panels show how the retained fraction of maximum permissible ABC changes as
the within-level fraction, $\Delta$, moves from the lower to upper part of each
risk interval. This illustration uses the four-category, three-level structure
used from 2023 onward; the applied table below uses the same three-level
structure but switches the number of categories by year.

```{r}
#| label: fig-thompson-heatmap
#| fig-cap: "Illustrative ABC retained under Thompson's probability link, summarized by the number of elevated categories and the common three-level risk score assigned to those categories. OFL is normalized to 1 and CV is set to 0.30."
#| fig-width: 8
#| fig-height: 5.5

thompson_p0 <- 0.49
thompson_nlev <- 3
thompson_ncat <- 4
thompson_cv <- 0.30
thompson_sigma <- sqrt(log1p(thompson_cv^2))

thompson_bounds_for_nlev <- function(nlev) {
  tibble(level = 1:nlev) |>
    mutate(
      p_upper = (level + nlev - 2) / (2 * (nlev - 1)),
      p_lower = if_else(level == 1, thompson_p0, lag(p_upper))
    )
}

thompson_bounds <- thompson_bounds_for_nlev(thompson_nlev)

level_to_p_cond <- function(level, delta = 0.5, nlev = thompson_nlev) {
  bounds <- thompson_bounds_for_nlev(nlev) |>
    filter(.data$level == .env$level)
  bounds$p_lower + delta * (bounds$p_upper - bounds$p_lower)
}

p_cond_to_marginal <- function(p_cond, ncat = thompson_ncat) {
  1 - ((1 - p_cond) / (1 - thompson_p0))^(1 / ncat)
}

joint_p <- function(pj) {
  1 - (1 - thompson_p0) * prod(1 - pj)
}

abc_retained <- function(p) {
  maxABC_pe_to_OFL <- qlnorm(
    p,
    meanlog = 0,
    sdlog = thompson_sigma
  )
  pmin(1, 1 / maxABC_pe_to_OFL)
}

thompson_heatmap <- crossing(
  n_high = 0:thompson_ncat,
  high_level = 1:thompson_nlev,
  delta = c(0.25, 0.50, 0.75)
) |>
  rowwise() |>
  mutate(
    levels = list(c(rep(high_level, n_high),
                    rep(1, thompson_ncat - n_high))),
    p_cond = list(map_dbl(levels, level_to_p_cond, delta = delta)),
    pj = list(map_dbl(p_cond, p_cond_to_marginal)),
    thompson_joint_p = joint_p(pj),
    ABC_over_maxABC = abc_retained(thompson_joint_p)
  ) |>
  ungroup()

ggplot(thompson_heatmap,
       aes(x = n_high, y = high_level, fill = ABC_over_maxABC)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(ABC_over_maxABC, 2)), size = 3) +
  facet_wrap(~ delta, labeller = label_both) +
  scale_fill_viridis_c(
    limits = c(0, 1),
    name = "ABC / maxABC"
  ) +
  scale_x_continuous(breaks = 0:thompson_ncat) +
  scale_y_continuous(breaks = 1:thompson_nlev) +
  labs(
    x = "Number of elevated-risk categories",
    y = "Three-level risk score of elevated categories"
  ) +
  theme_minimal(base_size = 12)
```

The first table below revisits the values Thompson applied conceptually. The
absolute value of $maxABC_{PE}$ depends on the OFL uncertainty distribution in
the assessment. For comparability, @tbl-thompson-levels normalizes OFL to 1,
uses a lognormal uncertainty distribution, sets $P_0 = 0.49$, and uses a CV of
0.30. The key column is therefore $maxABC_{PE}/OFL$, not an assessment-specific
tonnage. The CV can be changed in the code chunk if a specific stock's OFL
uncertainty is available.

```{r}
#| label: tbl-thompson-levels
#| tbl-cap: "Thompson risk-level probability intervals and probability-equivalent maxABC values under an illustrative lognormal OFL distribution."

thompson_bounds_midpoint <- thompson_bounds |>
  mutate(
    delta_midpoint = 0.5,
    p_cond_midpoint = p_lower + delta_midpoint * (p_upper - p_lower),
    maxABC_pe_to_OFL = qlnorm(
      p_cond_midpoint,
      meanlog = 0,
      sdlog = thompson_sigma
    )
  )

thompson_bounds_midpoint |>
  transmute(
    `Risk level` = level,
    `P lower` = round(p_lower, 3),
    `P upper` = round(p_upper, 3),
    `P midpoint` = round(p_cond_midpoint, 3),
    `maxABC_PE / OFL` = round(maxABC_pe_to_OFL, 3)
  ) |>
  kable()
```

To apply the same logic to the historical score combinations in this report,
@tbl-thompson-applied uses the year-appropriate number of categories, uses
three risk levels for all years, treats scores of 0 as normal level 1, and
sets the within-level fraction to $\Delta = 0.9$. This is not a reconstruction
of the SSC's
calculation, because the actual within-level $\Delta$ values and
stock-specific OFL uncertainty distributions are not in the summary table. It
is a transparent benchmark showing the probability-equivalent `maxABC` implied
by Thompson's mapping for the observed score combinations.

```{r}
#| label: tbl-thompson-applied
#| tbl-cap: "Illustrative application of Thompson's probability mapping to observed score combinations using three risk levels and within-level delta = 0.9. OFL is normalized to 1 and CV is set to 0.30."

thompson_applied_delta <- 0.9

year_settings <- function(year) {
  tibble(
    ncat = if_else(year == 2018, 3L, 4L),
    nlev = 3L,
    scoring_era = case_when(
      year == 2018 ~ "2018: 3 categories, 3 levels",
      TRUE ~ "2019+: 4 categories, 3 levels"
    )
  )
}

score_to_level <- function(score, nlev) {
  pmin(pmax(if_else(score == 0, 1, score), 1), nlev)
}

row_thompson <- function(A, P, E, F, year, delta = 0.5) {
  settings <- year_settings(year)
  scores <- if (settings$ncat == 3L) {
    c(A, P, E)
  } else {
    c(A, P, E, F)
  }
  levels <- score_to_level(scores, settings$nlev)
  p_cond <- map_dbl(levels, level_to_p_cond,
                    delta = delta, nlev = settings$nlev)
  pj <- map_dbl(p_cond, p_cond_to_marginal, ncat = settings$ncat)
  p <- joint_p(pj)
  maxABC_pe_to_OFL <- qlnorm(
    p,
    meanlog = 0,
    sdlog = thompson_sigma
  )
  tibble(
    ncat = settings$ncat,
    nlev = settings$nlev,
    scoring_era = settings$scoring_era,
    thompson_joint_p = p,
    maxABC_pe_to_OFL = maxABC_pe_to_OFL,
    thompson_abc_to_maxABC = pmin(1, 1 / maxABC_pe_to_OFL),
    thompson_pct_reduction = 100 * (1 - thompson_abc_to_maxABC)
  )
}

thompson_applied <- dat |>
  mutate(thompson = pmap(
    list(A, P, E, F, year),
    row_thompson,
    delta = thompson_applied_delta
  )) |>
  unnest(thompson)

thompson_applied |>
  transmute(
    Stock = stock,
    Year = year,
    A, P, E, F,
    `Scoring era` = scoring_era,
    ncat,
    nlev,
    `Observed SSC reduction (%)` = pct_red,
    `Joint P` = round(thompson_joint_p, 3),
    `maxABC_PE / OFL` = round(maxABC_pe_to_OFL, 3),
    `Thompson reduction (%)` = round(thompson_pct_reduction, 1)
  ) |>
  arrange(desc(`maxABC_PE / OFL`), Stock, Year) |>
  kable()
```

@fig-thompson-observed compares the observed SSC reductions with
three-level Thompson reductions across three high within-level values of
$\Delta$: 0.90, 0.95, and 0.99. Points
above the 1:1 line indicate cases where the illustrative Thompson calculation
would imply a larger reduction than the observed SSC reduction; points below
the line indicate cases where it would imply a smaller reduction.

```{r}
#| label: fig-thompson-observed
#| fig-cap: "Observed SSC reduction compared with the illustrative three-level Thompson reduction. Colors show years and shapes show within-level delta."
#| fig-width: 8
#| fig-height: 6

thompson_delta_grid <- dat |>
  crossing(delta = c(0.90, 0.95, 0.99)) |>
  mutate(thompson = pmap(
    list(A, P, E, F, year, delta),
    row_thompson
  )) |>
  unnest(thompson) |>
  mutate(delta = factor(delta))

ggplot(
  thompson_delta_grid,
  aes(
    x = pct_red,
    y = thompson_pct_reduction,
    color = factor(year),
    shape = delta
  )
) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey55") +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_viridis_d(name = "Year") +
  labs(
    x = "Observed SSC reduction from maxABC (%)",
    y = "Thompson reduction from maxABC (%)",
    color = "Year",
    shape = "Delta"
  ) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
  theme(legend.position = "right",
        legend.text = element_text(size = 8))
```

In plain language, Thompson's method turns the risk table into a probability
meter and then into a catch dial. The starting point is the assessment model's
baseline probability, $P_0$, that `maxABC` would exceed the true OFL. The risk
table then adds the concerns that are not already captured by the model:
assessment issues, population dynamics, ecosystem signals, and fishery
information. Each category is assigned a risk level, optionally fine-tuned
within that level with $\Delta$, and those category risks are combined into one
overall probability, $P$.

The decision rule is simple. If $P \leq 0.5$, the probability of overfishing is
treated as acceptable and ABC remains at `maxABC`. If $P > 0.5$, the
probability is too high and ABC is scaled down. Higher values of $P$ imply a
larger gap between `maxABC` and the probability-equivalent `maxABC`, so the
resulting ABC reduction gets larger.

The intuition is: few low-risk scores are a green light for full catch; some
moderate risks are a yellow light for a smaller cut; many high risks are a red
light for a larger cut. In one line, more risk means a higher chance of
overfishing, which means a lower allowed catch.

## Exploratory Figures

The exploratory displays in @fig-scatter-total, @fig-scatter-nelev,
@fig-facet-categories, and @fig-cor are intended to answer a simple question
before formal modeling begins: do the risk table scores contain visible signal
about the size of the final SSC reduction, and if so, what kind of signal is
it? Together these figures separate severity, breadth, category-specific
effects, and correlation structure.

### Reduction vs. total risk score

@fig-scatter-total motivates the most basic candidate model by asking
whether larger summed scores are generally associated with larger reductions.
If the total score were carrying most of the information in the risk table, a
clear monotonic relationship would be expected here.

```{r}
#| label: fig-scatter-total
#| fig-cap: "Percent reduction from maxABC vs. total risk score (sum of elevated category scores). Points are colored by stock. The dashed line is a simple linear fit."
#| fig-width: 8
#| fig-height: 5.5

ggplot(dat, aes(x = total_score, y = pct_red)) +
geom_smooth(method = "lm", formula = y ~ x,
              se = TRUE, color = "grey40", linetype = "dashed") +
  geom_point(aes(color = stock), size = 3, alpha = 0.85) +
  scale_color_brewer(palette = "Paired", name = "Stock") +
  labs(x = "Total risk score (sum of A + P + E + F)",
       y = "% reduction from maxABC") +
  theme(legend.position = "right",
        legend.text = element_text(size = 9))
```

The broad spread around the fitted line in @fig-scatter-total immediately
suggests that the total score alone will be an incomplete summary of the SSC
decision process, even if it captures some central tendency.

### Reduction vs. number of elevated categories

@fig-scatter-nelev shifts attention from severity to breadth. The goal
is to see whether reductions tend to increase as more categories are flagged,
even when the exact numerical scores differ across stocks and years.

```{r}
#| label: fig-scatter-nelev
#| fig-cap: "Percent reduction from maxABC vs. number of risk table categories scored above 1."
#| fig-width: 7
#| fig-height: 5

ggplot(dat, aes(x = factor(n_elevated), y = pct_red)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = "grey90") +
  geom_jitter(aes(color = stock), width = 0.12, size = 2.5, alpha = 0.85) +
  scale_color_brewer(palette = "Paired", name = "Stock") +
  labs(x = "Number of categories elevated (score > 1)",
       y = "% reduction from maxABC") +
  theme(legend.position = "right",
        legend.text = element_text(size = 9))
```

This view is useful because a stock with several modest concerns may be treated
differently from a stock with one isolated but severe concern, and
@fig-scatter-nelev makes that contrast easier to see than a summed score alone.

### Reduction by individual category score

@fig-facet-categories provides the main category-by-category context for
the later coefficient models. Rather than collapsing the table to a single
summary metric, it shows how the observed reductions vary with score level
within each of the four risk dimensions.

```{r}
#| label: fig-facet-categories
#| fig-cap: "Percent reduction from maxABC broken out by score level within each risk table category. Points are jittered horizontally."
#| fig-width: 9
#| fig-height: 5

dat |>
  pivot_longer(cols = c(A, P, E, F), names_to = "category", values_to = "score") |>
  mutate(category = recode(category,
    A = "Assessment (A)", P = "Population (P)",
    E = "Ecosystem (E)",  F = "Fishery (F)")) |>
  ggplot(aes(x = factor(score), y = pct_red)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7, color = "steelblue") +
  facet_wrap(~ category) +
  labs(x = "Risk score", y = "% reduction from maxABC")
```

This comparison helps assess whether some dimensions appear to matter more than
others. In particular, it previews whether population, ecosystem, or
assessment-related concerns tend to align more consistently with larger
reductions.

### Correlation matrix of predictors

@fig-cor is included as diagnostic context for model specification. The
purpose is not substantive on its own, but to show where predictors move
together and where they contain distinct information.

```{r}
#| label: fig-cor
#| fig-cap: "Pairwise correlation among risk score predictors and the percent reduction."
#| fig-width: 6
#| fig-height: 5

cor_mat <- dat |>
  select(A, P, E, F, total_score, pct_red) |>
  cor()

cor_long <- as.data.frame(as.table(cor_mat)) |>
  rename(var1 = Var1, var2 = Var2, r = Freq)

ggplot(cor_long, aes(var1, var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The correlation structure in @fig-cor informs whether a more aggregated model
is reasonable and helps interpret coefficient instability in models that
include all category scores simultaneously.

## Models

The frequentist models below provide progressively richer summaries of the
relationship between risk table scores and SSC reductions. A Bayesian
mixed-effects model is then added to account explicitly for repeated
observations from the same stock through partial pooling.

We consider four candidate models, ranging from simple to structured.

### Model 1: OLS on total score

A baseline linear model treating the sum of scores as the sole predictor.

```{r}
#| label: m1-fit

m1 <- lm(pct_red ~ total_score, data = dat)
summary(m1)
```

### Model 2: OLS with individual category scores

Allows each risk dimension to carry different weight.

```{r}
#| label: m2-fit

m2 <- lm(pct_red ~ A + P + E + F, data = dat)
summary(m2)
```

### Model 3: Beta regression on individual scores

Beta regression is the natural choice for a bounded proportional response.
We model the logit of the proportional reduction.

```{r}
#| label: m3-fit

m3 <- betareg(pct_red_prop ~ A + P + E + F, data = dat)
summary(m3)
```

### Model 4: Beta regression with total score + number elevated

A parsimonious alternative that separates severity (total score) from
breadth (how many categories are flagged).

```{r}
#| label: m4-fit

m4 <- betareg(pct_red_prop ~ total_score + n_elevated, data = dat)
summary(m4)
```

### Sensitivity: harmonized score scale

The main regression models above use the scores as recorded in the historical
risk table summary. Because the scoring scale changed in 2023, the sensitivity
models below harmonize 2023+ scores to the earlier four-level scale before
fitting the same broad model forms. Specifically, current level 2 is treated
as old level 3 and current level 3 is treated as old level 4. This adjustment
does not change whether a category is elevated, but it does change the
severity assigned to recent elevated scores.

```{r}
#| label: tbl-harmonized-sensitivity
#| tbl-cap: "Sensitivity comparison using raw scores and harmonized 2023+ scores."

m2_harmonized <- lm(
  pct_red ~ A_harmonized + P_harmonized + E_harmonized + F_harmonized,
  data = dat
)

m3_harmonized <- betareg(
  pct_red_prop ~ A_harmonized + P_harmonized +
    E_harmonized + F_harmonized,
  data = dat
)

m4_harmonized <- betareg(
  pct_red_prop ~ total_score_harmonized + n_elevated_harmonized,
  data = dat
)

harmonized_compare <- tribble(
  ~model, ~score_scale, ~description, ~AIC, ~R2,
  "M2", "Raw", "OLS ~ A + P + E + F",
    AIC(m2), summary(m2)$adj.r.squared,
  "M2h", "Harmonized", "OLS ~ A + P + E + F",
    AIC(m2_harmonized), summary(m2_harmonized)$adj.r.squared,
  "M3", "Raw", "Beta ~ A + P + E + F",
    AIC(m3), m3$pseudo.r.squared,
  "M3h", "Harmonized", "Beta ~ A + P + E + F",
    AIC(m3_harmonized), m3_harmonized$pseudo.r.squared,
  "M4", "Raw", "Beta ~ total + n_elevated",
    AIC(m4), m4$pseudo.r.squared,
  "M4h", "Harmonized", "Beta ~ total + n_elevated",
    AIC(m4_harmonized), m4_harmonized$pseudo.r.squared
)

harmonized_compare |>
  mutate(across(c(AIC, R2), ~ round(.x, 2))) |>
  kable(
    col.names = c(
      "Model", "Score scale", "Description", "AIC", "R² / pseudo-R²"
    )
  )
```

The harmonized fits should be read as a sensitivity analysis rather than a
replacement for the raw-score models. They ask whether the main conclusions
are sensitive to treating the 2023+ three-level scores as compressed versions
of the earlier four-level scores.

### Model 5: Bayesian beta mixed-effects model

The previous models treat stock-year observations as conditionally independent.
That is a strong assumption because repeated entries for the same stock may
share unmeasured characteristics that affect the SSC's choice of reduction.
To address that structure, the final model uses a beta likelihood with the same
four category scores as fixed effects and a stock-level random intercept.

For observation $i$ from stock $s[i]$, let $y_i$ denote the proportional
reduction from maxABC. The Bayesian model is:

$$
y_i \sim \text{Beta}(\mu_i \phi,\ (1 - \mu_i)\phi)
$$ {#eq-brms-response}

$$
\text{logit}(\mu_i) = \alpha + \beta_A A_i + \beta_P P_i + \beta_E E_i + \beta_F F_i + u_{s[i]}
$$ {#eq-brms-mean}

$$
u_s \sim \mathcal{N}(0, \sigma_{\text{stock}}), \qquad
\sigma_{\text{stock}} > 0
$$ {#eq-brms-ranef}

This formulation keeps the bounded-response structure of Model 3 while letting
each stock deviate from the overall intercept in a partially pooled way.
Weakly informative priors are used to stabilize estimation given the small
sample size.

```{r}
#| label: m5-fit
#| results: false

intercept_prior_mean <- qlogis(mean(dat$pct_red_prop))

priors_m5 <- c(
  prior_string(
    sprintf("normal(%0.6f, 1.5)", intercept_prior_mean),
    class = "Intercept"
  ),
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "phi")
)

m5 <- brm(
  formula = bf(pct_red_prop ~ A + P + E + F + (1 | stock)),
  data = dat,
  family = Beta(link = "logit"),
  prior = priors_m5,
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = min(2, parallel::detectCores()),
  seed = 1234,
  backend = "rstan",
  control = list(adapt_delta = 0.95),
  file = "m5_brms_fit",
  file_refit = "on_change",
  refresh = 0
)
```

@tbl-brms-fixed summarizes the posterior for the main coefficients and the
stock-level variance term. These estimates are easiest to interpret directionally:
positive values indicate larger expected reductions on the logit scale, after
accounting for the other scores and stock-to-stock heterogeneity.

```{r}
#| label: tbl-brms-fixed
#| tbl-cap: "Posterior summaries for the Bayesian beta mixed-effects model."

m5_post <- posterior_summary(
  m5,
  variable = c(
    "Intercept",
    "b_A", "b_P", "b_E", "b_F",
    "sd_stock__Intercept",
    "phi"
  ),
  probs = c(0.025, 0.975)
) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  mutate(
    parameter = recode(
      parameter,
      `Intercept` = "Intercept",
      `b_A` = "Assessment (A)",
      `b_P` = "Population (P)",
      `b_E` = "Ecosystem (E)",
      `b_F` = "Fishery (F)",
      `sd_stock__Intercept` = "Stock SD",
      `phi` = "Precision (phi)"
    )
  ) |>
  rename(
    Estimate = Estimate,
    `Est. Error` = Est.Error,
    `2.5%` = Q2.5,
    `97.5%` = Q97.5
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

kable(m5_post)
```

Because the Bayesian model includes stock-level partial pooling, it is not
strictly comparable to the AIC table below on the same scale. Instead, it is
best read as a complementary model that asks whether the main category effects
remain after allowing each stock to have its own baseline tendency toward
higher or lower reductions.

### Model comparison

@tbl-compare brings the four candidate models onto a common footing. The
motivation is pragmatic: rather than argue for a model based only on
interpretability, it compares fit statistics to show whether added structure is
buying meaningful explanatory value.

```{r}
#| label: tbl-compare
#| tbl-cap: "Model comparison. AIC and pseudo/adjusted R² where available."

compare <- tribble(
  ~model, ~description, ~AIC, ~R2,
  "M1", "OLS ~ total score",
    AIC(m1), summary(m1)$adj.r.squared,
  "M2", "OLS ~ A + P + E + F",
    AIC(m2), summary(m2)$adj.r.squared,
  "M3", "Beta ~ A + P + E + F",
    AIC(m3), m3$pseudo.r.squared,
  "M4", "Beta ~ total + n_elevated",
    AIC(m4), m4$pseudo.r.squared
)

compare |>
  mutate(across(c(AIC, R2), ~ round(.x, 2))) |>
  kable(col.names = c("Model", "Description", "AIC", "R² / pseudo-R²"))
```

Read @tbl-compare as a compact summary of the tradeoff between parsimony and
flexibility. It indicates whether it is more useful to model the four
categories separately or to collapse them into broader summaries like total
score and number elevated.

### Observed vs. fitted

@fig-obs-fit provides the most direct visual model check in the report.
Its purpose is to show how closely each candidate model reproduces the observed
reductions across the full sample, and whether any model systematically
underpredicts or overpredicts certain cases.

```{r}
#| label: fig-obs-fit
#| fig-cap: "Observed vs. fitted reductions for the five candidate models. The 1:1 line is shown for reference."
#| fig-width: 9
#| fig-height: 7

fits <- dat |>
  mutate(
    M1 = fitted(m1),
    M2 = fitted(m2),
    M3 = fitted(m3) * 100,
    M4 = fitted(m4) * 100,
    M5 = fitted(m5, summary = TRUE)[, "Estimate"] * 100
  ) |>
  pivot_longer(M1:M5, names_to = "model", values_to = "fitted") |>
  mutate(model = recode(model,
    M1 = "M1: OLS ~ total",
    M2 = "M2: OLS ~ A+P+E+F",
    M3 = "M3: Beta ~ A+P+E+F",
    M4 = "M4: Beta ~ total+n_elev",
    M5 = "M5: Bayesian beta + stock RE"))

ggplot(fits, aes(x = fitted, y = pct_red)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = stock), size = 2, alpha = 0.8) +
  facet_wrap(~ model, scales = "free_x") +
  scale_color_brewer(palette = "Paired", name = "Stock") +
  labs(x = "Fitted % reduction", y = "Observed % reduction") +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

Because the 1:1 line is drawn explicitly in @fig-obs-fit, departures from that
line make it easier to identify which observations are structurally hard to
explain regardless of model choice.

### Coefficient effects (beta regression)

@fig-coefs translates the preferred category-based beta regression into
an interpretable effect display. The main motivation is to show direction,
magnitude, and uncertainty for each category after accounting for the others in
the same model.

```{r}
#| label: fig-coefs
#| fig-cap: "Estimated marginal effect of each risk score category on percent reduction (Model 3, beta regression). Points show the coefficient estimate on the logit scale; bars are 95% confidence intervals."
#| fig-width: 6
#| fig-height: 4

coef_tbl <- broom::tidy(m3, conf.int = TRUE) |>
  filter(component == "mean", term != "(Intercept)") |>
  mutate(term = recode(term,
    A = "Assessment (A)", P = "Population (P)",
    E = "Ecosystem (E)",  F = "Fishery (F)"))

ggplot(coef_tbl, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  size = 0.8, color = "steelblue") +
  labs(x = "Coefficient (logit scale)", y = NULL)
```

In that sense, @fig-coefs is the clearest summary of which risk dimensions are
most strongly associated with larger reductions once the overlapping signal
among categories is taken into account.

## Residual diagnostics

@fig-resid addresses model adequacy rather than fishery interpretation.
The intent is to check whether the beta regression leaves obvious structure in
the residuals or shows departures large enough to undermine the main inferences.

```{r}
#| label: fig-resid
#| fig-cap: "Residual diagnostics for Model 3 (beta regression with individual category scores)."
#| fig-width: 9
#| fig-height: 4

p_resid <- dat |>
  mutate(fitted = fitted(m3),
         resid  = residuals(m3, type = "sweighted2")) |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 2, alpha = 0.7) +
  labs(x = "Fitted value", y = "Weighted residual",
       subtitle = "Residuals vs. fitted")

p_qq <- dat |>
  mutate(resid = residuals(m3, type = "sweighted2")) |>
  ggplot(aes(sample = resid)) +
  stat_qq(size = 2, alpha = 0.7) +
  stat_qq_line(color = "steelblue") +
  labs(x = "Theoretical quantile", y = "Sample quantile",
       subtitle = "Normal Q-Q plot")

p_resid + p_qq
```

These diagnostics do not need to look perfect to be useful, but @fig-resid
helps distinguish a modestly noisy descriptive model from one that is plainly
misspecified.

## Discussion

Several patterns emerge:

1. **Population dynamics (P) scores are the strongest individual predictor**
   of larger reductions, consistent with the SSC weighting biomass trajectory
   and recruitment concerns heavily.
   That signal persists in the Bayesian mixed-effects fit in @tbl-brms-fixed,
   although the posterior intervals remain fairly wide.

2. **Total score is a weak overall predictor.** The SSC clearly applies
   considerable judgment beyond what the risk scores encode—stock identity,
   economic context, and the specific nature of each concern all matter.
   This is consistent with the dispersion visible in @fig-scatter-total and
   the comparison between aggregated and category-level models in @tbl-compare.

3. **Outliers are informative.** EBS Pollock 2022 (A-2 only, 43% reduction)
   and EBS Pollock 2021 (A-2, E-2, F-2, only 11%) illustrate that
   stock-specific context can dominate the risk table signal.
   Those difficult cases are also evident in @fig-obs-fit, where even better
   performing models do not fully recover every stock-year outcome.

4. **Sample size limits model complexity.** With *n* = 21 observations across
   13 stocks, more structured models are feasible but still uncertain.
   The Bayesian mixed-effects model in @tbl-brms-fixed helps by partially
   pooling repeated stock observations, yet the intervals remain wide enough
   that the fitted relationships should still be treated cautiously.

5. **The risk table is an input to, not a determinant of, the SSC decision.**
   Any predictive model should be understood as describing central tendency
   rather than prescribing outcomes.

## Report context and interpretation

The broader 2024 update emphasized that reductions from maximum permissible
ABC have occurred for a limited set of stocks over the lifetime of the risk
table process. Those stocks include one Alaska-wide stock (sablefish), eight
stocks from the BSAI, and five from the GOA. In most cases where authors
recommended a reduction, the SSC agreed, but there were also important
exceptions where the SSC either imposed a reduction that authors had not
proposed or declined to adopt a proposed reduction.

That pattern aligns with the fitted results here. The scores contain real
signal, especially for population and broader multi-category concern, but they
do not encode the full decision process. The SSC appears to use the risk table
as a structured summary of uncertainty and concern, then combine it with
stock-specific considerations, model behavior, and management judgment when
setting the final buffer below maxABC.

For that reason, the models in this document are best used as exploratory
benchmarks. They help summarize what kinds of score combinations have
historically coincided with larger reductions, but they should not be treated
as a mechanical rule for future assessments operating under a still-evolving
risk table framework.
