---
title: "Modeling SSC Reductions from maxABC Using Risk Table Scores"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    lightbox: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    self-contained: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(betareg)
library(brms)
library(knitr)
library(patchwork)

theme_set(theme_minimal(base_size = 13))
options(mc.cores = min(2, parallel::detectCores()))
```

## Summary

This report looks at whether groundfish risk table scores help explain when
the Scientific and Statistical Committee (SSC) recommends setting ABC below
the maximum permissible ABC. In simple terms, it asks whether higher or more
widespread concerns in the risk table tend to line up with larger reductions.

The answer is yes, but only partly. Stocks with more elevated concerns often
do receive larger reductions, and population dynamics concerns appear to be
especially informative. At the same time, the relationship is not tight enough
to treat the risk table as a formula. Similar score patterns can still lead to
different reductions because the SSC also considers stock-specific context,
model behavior, and management judgment.

The results should therefore be read as descriptive rather than prescriptive.
They summarize how risk table scores and SSC reductions have lined up in the
historical record, but they do not provide a mechanical rule for future
decisions.

## Introduction

The North Pacific Fishery Management Council's Scientific and Statistical
Committee (SSC) may set ABCs below the maximum permissible ABC (maxABC)
when the assessment risk table identifies elevated concerns. The risk table
scores four categories—Assessment-related (A), Population dynamics (P),
Ecosystem (E), and Fishery-informed (F)—on a scale where scores >1 indicate
concern.

This analysis sits within the broader context of the *Groundfish Risk Table
Report: 2024 Update*, which collated risk table information for Alaska
groundfish stocks in the Bering Sea and Aleutian Islands (BSAI) and the Gulf
of Alaska (GOA) from 2018 onward. Risk table implementation changed
substantially over that period. The 2018 trial year included only a small
number of completed tables and only three categories. Fishery performance was
added in 2019, and in response to SSC direction the scoring system was reduced
from four levels to three beginning in 2023. Category descriptions were then
revised again for 2024 to better align with the intent of the original table.
Those shifts make direct comparisons across years imperfect, especially when
contrasting the 2018 trial year and the reduced-level years from 2023 onward
with the 2019-2022 period.

An AFSC working group convened in early 2024 to review all groundfish risk
table scores to date and consolidate them into a single living reference.
That effort documented annual category scores, the maximum score across
categories, author-proposed reductions, SSC-recommended reductions, the
methods used to derive those reductions, and cases where the author, Plan
Team, and SSC differed. The working group used that summary both to support
stock assessment authors and to identify common patterns in how risk tables
have been translated into ABC reductions over time.

Across 2018-2024, a total of 136 risk tables were developed. Of those, 77 had
at least one elevated score, 59 had no elevated scores, 25 included an author
recommendation to reduce from maximum ABC, and 28 resulted in a final
reduction following Plan Team and SSC review. Even with that history, the
overall number of reductions remains small relative to the number of risk
tables produced, which reinforces that risk table scores are informative but
not determinative on their own.

This document narrows that broader report into a modeling exercise. It focuses
on the relationship between risk table scores and the magnitude of the SSC's
recommended reduction from maxABC, and evaluates several candidate models for
describing those reductions.

## Background

### How the risk table framework evolved

Several design changes matter for interpretation of any time series built from
these scores:

1. In 2018, risk tables were used on a limited trial basis and included only
   assessment, population dynamics, and ecosystem categories.
2. In 2019, fishery performance was added as a fourth category.
3. In 2023, the four scoring levels were reduced to three: normal, increased,
   and extreme. Under that reduced system, current levels 2 and 3 are roughly
   comparable to the earlier levels 3 and 4.
4. In 2024, the text defining the categories was updated after the SSC asked
   for descriptions that better covered the full intended range of conditions.

These changes mean the raw scores are not fully stationary through time. Any
model fit to the combined record should therefore be treated as descriptive of
historical SSC behavior rather than as a stable calibration of a fixed scoring
system.

### How reductions have been derived in practice

The 2024 summary report documented several approaches authors and reviewers
have used when moving from elevated scores to a reduction from maxABC. Those
approaches include:

1. Deferring to the SSC to specify a reduction if one was judged warranted.
2. Applying tier-based advice developed for one stock status category to
   another, such as Tier 3-style advice used for Tier 1 stocks.
3. Using prior projection results when a previous scenario better captured the
   concern driving the reduction.
4. Applying probability-based methods such as the Thompson framework for
   translating elevated risk into a buffer below maxABC.
5. Selecting an ABC level intended to keep the stock above a biological
   threshold such as B20%.
6. Using an alternative model configuration with better retrospective
   performance or lower OFL and ABC values.
7. Taking a stair-step approach between the current-year estimate and a
   previous projection.
8. Adjusting recent recruitment assumptions or the reference period used in
   projections.

That diversity in methods is important for interpreting the results below. The
same nominal risk score can map to different reductions depending on stock
status, model structure, and the rationale judged most relevant by the SSC.

## Data

The data below come from SSC-recommended reductions only (not Plan Team or
author recommendations). Rollover years and records without a quantified
reduction are excluded. Blank years in the broader report generally correspond
to off-cycle assessments with no new risk table, although a reduction may
still carry over from an earlier decision. The analysis dataset below is
therefore a filtered subset of the full living summary rather than a complete
reproduction of every annual stock entry.

@tbl-data is the core analytical dataset used throughout the document.
It provides the stock-year observations, the category-level elevated scores
used as predictors, and the SSC reduction used as the response variable. The
derived columns summarize two distinct ideas that recur in the figures and
models below: overall severity through the total score, and breadth of concern
through the number of elevated categories.

```{r}
#| label: tbl-data
#| tbl-cap: "Risk table scores and SSC-recommended percent reductions from maxABC. Scores of 0 indicate the category was not elevated (score ≤ 1)."

dat <- tribble(
  ~stock,                            ~year, ~A, ~P, ~E, ~F, ~pct_red,
  "EBS Pollock",                      2018,  0,  2,  2,  0,  30,
  "EBS Pollock",                      2019,  0,  2,  2,  2,  43,
  "EBS Pollock",                      2020,  0,  0,  2,  2,  30,
  "EBS Pollock",                      2021,  2,  0,  2,  2,  11,
  "EBS Pollock",                      2022,  2,  0,  0,  0,  43,
  "EBS Pollock",                      2023,  0,  0,  2,  0,  18,
  "EBS Pacific cod",                  2019,  0,  0,  2,  0,   3,
  "AI Pacific cod",                   2023,  0,  2,  2,  0,  10,
  "AK Sablefish",                     2018,  2,  4,  2,  0,  45,
  "AK Sablefish",                     2019,  2,  3,  2,  3,  50,
  "AK Sablefish",                     2020,  3,  3,  2,  3,  44,
  "BSAI Yellowfin sole",              2021,  0,  2,  2,  0,  24,
  "BSAI Greenland turbot",            2024,  3,  2,  2,  0,  25,
  "BSAI Northern rock sole",          2022,  3,  0,  0,  0,  23,
  "BSAI Blackspotted/Rougheye RF",   2022,  3,  2,  0,  2,  11,
  "BSAI Sharks",                      2022,  3,  2,  0,  0,  13,
  "GOA Pollock",                      2023,  2,  0,  0,  0,  18,
  "GOA Dusky Rockfish",               2020,  2,  0,  0,  0,  24,
  "GOA Rougheye/Blackspotted RF",     2023,  2,  2,  0,  0,  20,
  "GOA Demersal shelf rockfish",      2021,  2,  2,  0,  2,  22,
  "GOA Demersal shelf rockfish",      2022,  2,  2,  0,  2,  15
)

dat <- dat |>
  mutate(
    total_score  = A + P + E + F,
    n_elevated   = (A > 0) + (P > 0) + (E > 0) + (F > 0),
    pct_red_prop = pct_red / 100
  )

dat |>
  select(stock, year, A, P, E, F, total_score, n_elevated, pct_red) |>
  kable(col.names = c("Stock", "Year", "A", "P", "E", "F",
                       "Total", "# Elevated", "% Reduction"))
```

## Exploratory Figures

The exploratory displays in @fig-scatter-total, @fig-scatter-nelev,
@fig-facet-categories, and @fig-cor are intended to answer a simple question
before formal modeling begins: do the risk table scores contain visible signal
about the size of the final SSC reduction, and if so, what kind of signal is
it? Together these figures separate severity, breadth, category-specific
effects, and correlation structure.

### Reduction vs. total risk score

@fig-scatter-total motivates the most basic candidate model by asking
whether larger summed scores are generally associated with larger reductions.
If the total score were carrying most of the information in the risk table, a
clear monotonic relationship would be expected here.

```{r}
#| label: fig-scatter-total
#| fig-cap: "Percent reduction from maxABC vs. total risk score (sum of elevated category scores). Points are colored by stock. The dashed line is a simple linear fit."
#| fig-width: 8
#| fig-height: 5.5

ggplot(dat, aes(x = total_score, y = pct_red)) +
geom_smooth(method = "lm", formula = y ~ x,
              se = TRUE, color = "grey40", linetype = "dashed") +
  geom_point(aes(color = stock), size = 3, alpha = 0.85) +
  scale_color_brewer(palette = "Paired", name = "Stock") +
  labs(x = "Total risk score (sum of A + P + E + F)",
       y = "% reduction from maxABC") +
  theme(legend.position = "right",
        legend.text = element_text(size = 9))
```

The broad spread around the fitted line in @fig-scatter-total immediately
suggests that the total score alone will be an incomplete summary of the SSC
decision process, even if it captures some central tendency.

### Reduction vs. number of elevated categories

@fig-scatter-nelev shifts attention from severity to breadth. The goal
is to see whether reductions tend to increase as more categories are flagged,
even when the exact numerical scores differ across stocks and years.

```{r}
#| label: fig-scatter-nelev
#| fig-cap: "Percent reduction from maxABC vs. number of risk table categories scored above 1."
#| fig-width: 7
#| fig-height: 5

ggplot(dat, aes(x = factor(n_elevated), y = pct_red)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = "grey90") +
  geom_jitter(aes(color = stock), width = 0.12, size = 2.5, alpha = 0.85) +
  scale_color_brewer(palette = "Paired", name = "Stock") +
  labs(x = "Number of categories elevated (score > 1)",
       y = "% reduction from maxABC") +
  theme(legend.position = "right",
        legend.text = element_text(size = 9))
```

This view is useful because a stock with several modest concerns may be treated
differently from a stock with one isolated but severe concern, and
@fig-scatter-nelev makes that contrast easier to see than a summed score alone.

### Reduction by individual category score

@fig-facet-categories provides the main category-by-category context for
the later coefficient models. Rather than collapsing the table to a single
summary metric, it shows how the observed reductions vary with score level
within each of the four risk dimensions.

```{r}
#| label: fig-facet-categories
#| fig-cap: "Percent reduction from maxABC broken out by score level within each risk table category. Points are jittered horizontally."
#| fig-width: 9
#| fig-height: 5

dat |>
  pivot_longer(cols = c(A, P, E, F), names_to = "category", values_to = "score") |>
  mutate(category = recode(category,
    A = "Assessment (A)", P = "Population (P)",
    E = "Ecosystem (E)",  F = "Fishery (F)")) |>
  ggplot(aes(x = factor(score), y = pct_red)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7, color = "steelblue") +
  facet_wrap(~ category) +
  labs(x = "Risk score", y = "% reduction from maxABC")
```

This comparison helps assess whether some dimensions appear to matter more than
others. In particular, it previews whether population, ecosystem, or
assessment-related concerns tend to align more consistently with larger
reductions.

### Correlation matrix of predictors

@fig-cor is included as diagnostic context for model specification. The
purpose is not substantive on its own, but to show where predictors move
together and where they contain distinct information.

```{r}
#| label: fig-cor
#| fig-cap: "Pairwise correlation among risk score predictors and the percent reduction."
#| fig-width: 6
#| fig-height: 5

cor_mat <- dat |>
  select(A, P, E, F, total_score, pct_red) |>
  cor()

cor_long <- as.data.frame(as.table(cor_mat)) |>
  rename(var1 = Var1, var2 = Var2, r = Freq)

ggplot(cor_long, aes(var1, var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The correlation structure in @fig-cor informs whether a more aggregated model
is reasonable and helps interpret coefficient instability in models that
include all category scores simultaneously.

## Models

The frequentist models below provide progressively richer summaries of the
relationship between risk table scores and SSC reductions. A Bayesian
mixed-effects model is then added to account explicitly for repeated
observations from the same stock through partial pooling.

We consider four candidate models, ranging from simple to structured.

### Model 1: OLS on total score

A baseline linear model treating the sum of scores as the sole predictor.

```{r}
#| label: m1-fit

m1 <- lm(pct_red ~ total_score, data = dat)
summary(m1)
```

### Model 2: OLS with individual category scores

Allows each risk dimension to carry different weight.

```{r}
#| label: m2-fit

m2 <- lm(pct_red ~ A + P + E + F, data = dat)
summary(m2)
```

### Model 3: Beta regression on individual scores

Beta regression is the natural choice for a bounded proportional response.
We model the logit of the proportional reduction.

```{r}
#| label: m3-fit

m3 <- betareg(pct_red_prop ~ A + P + E + F, data = dat)
summary(m3)
```

### Model 4: Beta regression with total score + number elevated

A parsimonious alternative that separates severity (total score) from
breadth (how many categories are flagged).

```{r}
#| label: m4-fit

m4 <- betareg(pct_red_prop ~ total_score + n_elevated, data = dat)
summary(m4)
```

### Model 5: Bayesian beta mixed-effects model

The previous models treat stock-year observations as conditionally independent.
That is a strong assumption because repeated entries for the same stock may
share unmeasured characteristics that affect the SSC's choice of reduction.
To address that structure, the final model uses a beta likelihood with the same
four category scores as fixed effects and a stock-level random intercept.

For observation $i$ from stock $s[i]$, let $y_i$ denote the proportional
reduction from maxABC. The Bayesian model is:

$$
y_i \sim \text{Beta}(\mu_i \phi,\ (1 - \mu_i)\phi)
$$ {#eq-brms-response}

$$
\text{logit}(\mu_i) = \alpha + \beta_A A_i + \beta_P P_i + \beta_E E_i + \beta_F F_i + u_{s[i]}
$$ {#eq-brms-mean}

$$
u_s \sim \mathcal{N}(0, \sigma_{\text{stock}}), \qquad
\sigma_{\text{stock}} > 0
$$ {#eq-brms-ranef}

This formulation keeps the bounded-response structure of Model 3 while letting
each stock deviate from the overall intercept in a partially pooled way.
Weakly informative priors are used to stabilize estimation given the small
sample size.

```{r}
#| label: m5-fit
#| results: false

intercept_prior_mean <- qlogis(mean(dat$pct_red_prop))

priors_m5 <- c(
  prior_string(
    sprintf("normal(%0.6f, 1.5)", intercept_prior_mean),
    class = "Intercept"
  ),
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "phi")
)

m5 <- brm(
  formula = bf(pct_red_prop ~ A + P + E + F + (1 | stock)),
  data = dat,
  family = Beta(link = "logit"),
  prior = priors_m5,
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = min(2, parallel::detectCores()),
  seed = 1234,
  backend = "rstan",
  control = list(adapt_delta = 0.95),
  file = "m5_brms_fit",
  file_refit = "on_change",
  refresh = 0
)
```

@tbl-brms-fixed summarizes the posterior for the main coefficients and the
stock-level variance term. These estimates are easiest to interpret directionally:
positive values indicate larger expected reductions on the logit scale, after
accounting for the other scores and stock-to-stock heterogeneity.

```{r}
#| label: tbl-brms-fixed
#| tbl-cap: "Posterior summaries for the Bayesian beta mixed-effects model."

m5_post <- posterior_summary(
  m5,
  variable = c(
    "Intercept",
    "b_A", "b_P", "b_E", "b_F",
    "sd_stock__Intercept",
    "phi"
  ),
  probs = c(0.025, 0.975)
) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  mutate(
    parameter = recode(
      parameter,
      `Intercept` = "Intercept",
      `b_A` = "Assessment (A)",
      `b_P` = "Population (P)",
      `b_E` = "Ecosystem (E)",
      `b_F` = "Fishery (F)",
      `sd_stock__Intercept` = "Stock SD",
      `phi` = "Precision (phi)"
    )
  ) |>
  rename(
    Estimate = Estimate,
    `Est. Error` = Est.Error,
    `2.5%` = Q2.5,
    `97.5%` = Q97.5
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

kable(m5_post)
```

Because the Bayesian model includes stock-level partial pooling, it is not
strictly comparable to the AIC table below on the same scale. Instead, it is
best read as a complementary model that asks whether the main category effects
remain after allowing each stock to have its own baseline tendency toward
higher or lower reductions.

### Model comparison

@tbl-compare brings the four candidate models onto a common footing. The
motivation is pragmatic: rather than argue for a model based only on
interpretability, it compares fit statistics to show whether added structure is
buying meaningful explanatory value.

```{r}
#| label: tbl-compare
#| tbl-cap: "Model comparison. AIC and pseudo/adjusted R² where available."

compare <- tribble(
  ~model, ~description, ~AIC, ~R2,
  "M1", "OLS ~ total score",
    AIC(m1), summary(m1)$adj.r.squared,
  "M2", "OLS ~ A + P + E + F",
    AIC(m2), summary(m2)$adj.r.squared,
  "M3", "Beta ~ A + P + E + F",
    AIC(m3), m3$pseudo.r.squared,
  "M4", "Beta ~ total + n_elevated",
    AIC(m4), m4$pseudo.r.squared
)

compare |>
  mutate(across(c(AIC, R2), ~ round(.x, 2))) |>
  kable(col.names = c("Model", "Description", "AIC", "R² / pseudo-R²"))
```

Read @tbl-compare as a compact summary of the tradeoff between parsimony and
flexibility. It indicates whether it is more useful to model the four
categories separately or to collapse them into broader summaries like total
score and number elevated.

### Observed vs. fitted

@fig-obs-fit provides the most direct visual model check in the report.
Its purpose is to show how closely each candidate model reproduces the observed
reductions across the full sample, and whether any model systematically
underpredicts or overpredicts certain cases.

```{r}
#| label: fig-obs-fit
#| fig-cap: "Observed vs. fitted reductions for the five candidate models. The 1:1 line is shown for reference."
#| fig-width: 9
#| fig-height: 7

fits <- dat |>
  mutate(
    M1 = fitted(m1),
    M2 = fitted(m2),
    M3 = fitted(m3) * 100,
    M4 = fitted(m4) * 100,
    M5 = fitted(m5, summary = TRUE)[, "Estimate"] * 100
  ) |>
  pivot_longer(M1:M5, names_to = "model", values_to = "fitted") |>
  mutate(model = recode(model,
    M1 = "M1: OLS ~ total",
    M2 = "M2: OLS ~ A+P+E+F",
    M3 = "M3: Beta ~ A+P+E+F",
    M4 = "M4: Beta ~ total+n_elev",
    M5 = "M5: Bayesian beta + stock RE"))

ggplot(fits, aes(x = fitted, y = pct_red)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = stock), size = 2, alpha = 0.8) +
  facet_wrap(~ model, scales = "free_x") +
  scale_color_brewer(palette = "Paired", name = "Stock") +
  labs(x = "Fitted % reduction", y = "Observed % reduction") +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

Because the 1:1 line is drawn explicitly in @fig-obs-fit, departures from that
line make it easier to identify which observations are structurally hard to
explain regardless of model choice.

### Coefficient effects (beta regression)

@fig-coefs translates the preferred category-based beta regression into
an interpretable effect display. The main motivation is to show direction,
magnitude, and uncertainty for each category after accounting for the others in
the same model.

```{r}
#| label: fig-coefs
#| fig-cap: "Estimated marginal effect of each risk score category on percent reduction (Model 3, beta regression). Points show the coefficient estimate on the logit scale; bars are 95% confidence intervals."
#| fig-width: 6
#| fig-height: 4

coef_tbl <- broom::tidy(m3, conf.int = TRUE) |>
  filter(component == "mean", term != "(Intercept)") |>
  mutate(term = recode(term,
    A = "Assessment (A)", P = "Population (P)",
    E = "Ecosystem (E)",  F = "Fishery (F)"))

ggplot(coef_tbl, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  size = 0.8, color = "steelblue") +
  labs(x = "Coefficient (logit scale)", y = NULL)
```

In that sense, @fig-coefs is the clearest summary of which risk dimensions are
most strongly associated with larger reductions once the overlapping signal
among categories is taken into account.

## Residual diagnostics

@fig-resid addresses model adequacy rather than fishery interpretation.
The intent is to check whether the beta regression leaves obvious structure in
the residuals or shows departures large enough to undermine the main inferences.

```{r}
#| label: fig-resid
#| fig-cap: "Residual diagnostics for Model 3 (beta regression with individual category scores)."
#| fig-width: 9
#| fig-height: 4

p_resid <- dat |>
  mutate(fitted = fitted(m3),
         resid  = residuals(m3, type = "sweighted2")) |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 2, alpha = 0.7) +
  labs(x = "Fitted value", y = "Weighted residual",
       subtitle = "Residuals vs. fitted")

p_qq <- dat |>
  mutate(resid = residuals(m3, type = "sweighted2")) |>
  ggplot(aes(sample = resid)) +
  stat_qq(size = 2, alpha = 0.7) +
  stat_qq_line(color = "steelblue") +
  labs(x = "Theoretical quantile", y = "Sample quantile",
       subtitle = "Normal Q-Q plot")

p_resid + p_qq
```

These diagnostics do not need to look perfect to be useful, but @fig-resid
helps distinguish a modestly noisy descriptive model from one that is plainly
misspecified.

## Discussion

Several patterns emerge:

1. **Population dynamics (P) scores are the strongest individual predictor**
   of larger reductions, consistent with the SSC weighting biomass trajectory
   and recruitment concerns heavily.
   That signal persists in the Bayesian mixed-effects fit in @tbl-brms-fixed,
   although the posterior intervals remain fairly wide.

2. **Total score is a weak overall predictor.** The SSC clearly applies
   considerable judgment beyond what the risk scores encode—stock identity,
   economic context, and the specific nature of each concern all matter.
   This is consistent with the dispersion visible in @fig-scatter-total and
   the comparison between aggregated and category-level models in @tbl-compare.

3. **Outliers are informative.** EBS Pollock 2022 (A-2 only, 43% reduction)
   and EBS Pollock 2021 (A-2, E-2, F-2, only 11%) illustrate that
   stock-specific context can dominate the risk table signal.
   Those difficult cases are also evident in @fig-obs-fit, where even better
   performing models do not fully recover every stock-year outcome.

4. **Sample size limits model complexity.** With *n* = 21 observations across
   13 stocks, more structured models are feasible but still uncertain.
   The Bayesian mixed-effects model in @tbl-brms-fixed helps by partially
   pooling repeated stock observations, yet the intervals remain wide enough
   that the fitted relationships should still be treated cautiously.

5. **The risk table is an input to, not a determinant of, the SSC decision.**
   Any predictive model should be understood as describing central tendency
   rather than prescribing outcomes.

## Report context and interpretation

The broader 2024 update emphasized that reductions from maximum permissible
ABC have occurred for a limited set of stocks over the lifetime of the risk
table process. Those stocks include one Alaska-wide stock (sablefish), eight
stocks from the BSAI, and five from the GOA. In most cases where authors
recommended a reduction, the SSC agreed, but there were also important
exceptions where the SSC either imposed a reduction that authors had not
proposed or declined to adopt a proposed reduction.

That pattern aligns with the fitted results here. The scores contain real
signal, especially for population and broader multi-category concern, but they
do not encode the full decision process. The SSC appears to use the risk table
as a structured summary of uncertainty and concern, then combine it with
stock-specific considerations, model behavior, and management judgment when
setting the final buffer below maxABC.

For that reason, the models in this document are best used as exploratory
benchmarks. They help summarize what kinds of score combinations have
historically coincided with larger reductions, but they should not be treated
as a mechanical rule for future assessments operating under a still-evolving
risk table framework.
